---
title: "Abolish the Monarchy"
author: "Fraass"
date: "3/29/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This file walks through the analysis of the morphometric data from the entire range of planktic foraminifera. It was originally a part of a back-and-forth conversation between C. Lowery and myself (A. Fraass) as a part of writing the Lowery and Fraass 2019 paper. It has also been used as a part of writing the sequel paper Fraass, Lam, Lowery, and Jamson (Abolish the Monarchy) 

```{r loading}
read.csv(file="/MasterList (vstat 1.05).csv",header=TRUE)->foram.dataframe
foram.dataframe->morph

"%w/o%" <- function(x, y) x[!x %in% y] #--  x without y

exclude<-c(4)
morph[1:length(morph[,1] %w/o% exclude),]->morph
```
This adds the files, excluding the one that is missing data.

```{r custom functions, include=FALSE}
time.mean<-function(data,origin,extin,resolution=0.1){
  max.time<-max(origin)
  min.time<-min(extin)
  seq(to=max.time,from=min.time,by=resolution)->t
  cbind(t,rep(NA,times=length(t)))->output
  for(i in 1:length(t)){
    mean(data[which(origin > t[i] & extin < t[i])],na.rm=T)->output[i,2]
  }
  return(output)
}
time.div<-function(data,origin,extin,resolution=0.1){
  max.time<-max(origin)
  min.time<-min(extin)
  seq(to=max.time,from=min.time,by=resolution)->t
  cbind(t,rep(NA,times=length(t)))->output
  for(i in 1:length(t)){
    length(test.complex.index[which(origin > t[i] & extin < t[i])])->output[i,2]
    output[i,2]-1->output[i,2]
    }
  return(output)
}
time.median<-function(data,origin,extin,resolution=0.1){
  max.time<-max(origin)
  min.time<-min(extin)
  seq(to=max.time,from=min.time,by=resolution)->t
  cbind(t,rep(NA,times=length(t)))->output
  for(i in 1:length(t)){
    median(data[which(origin > t[i] & extin < t[i])],na.rm=T)->output[i,2]
  }
  return(output)
}
time.var<-function(data,origin,extin,resolution=0.1){
  max.time<-max(origin)
  min.time<-min(extin)
  seq(to=max.time,from=min.time,by=resolution)->t
  cbind(t,rep(NA,times=length(t)))->output
  for(i in 1:length(t)){
    var(data[which(origin > t[i] & extin < t[i])],na.rm=T)->output[i,2]
  }
  return(output)
}
library(vegan)
library(RColorBrewer)
library(vioplot)
range.line.plot<-function(var.A,y.axis.label,family.ID,origin,extin){
  #this was updated in 2022 because of change to the way that levels were functioning. Now instead of running on the levels
    #there's a unique() function used on the family.ID. levels() and droplevels() were no longer functioning.
  
  #Setting the spacer to make the vioplots visible
  spacer<-4.5
  #setting colors
  #these are based on the colors set in the other function (vioplot.pca)
  c(brewer.pal(12,'Set3'),brewer.pal(8,"Set2"),brewer.pal(9,'Set1'),'black','grey','black','brown')->col.temp
  fam.col<-rep(NA,length(family.ID))
  #droplevels(family.ID)->family.ID
  col.code<-rep(NA,times=length(levels(family.ID)))
  for(i in 1:length(unique(family.ID))){
    unique(family.ID)[i]->temp.ID
    if(temp.ID == "Candeinidae"){col.code[i]<-1}
    if(temp.ID == "Cassigerinellidae"){col.code[i]<-3}
    if(temp.ID == "Chiloguembelinidae"){col.code[i]<-4}
    if(temp.ID == "Favusellidae"){col.code[i]<-5}
    if(temp.ID == "Globigerinelloididae"){col.code[i]<-6}
    if(temp.ID == "Globigerinidae"){col.code[i]<-7}
    if(temp.ID == "Globoquadrinidae"){col.code[i]<-8}
    if(temp.ID == "Globorotaliidae"){col.code[i]<-9}
    if(temp.ID == "Globotruncanidae"){col.code[i]<-10}
    if(temp.ID == "Globuligerinidae"){col.code[i]<-11}
    if(temp.ID == "Guembelitriidae"){col.code[i]<-13}
    if(temp.ID == "Hantkeninidae"){col.code[i]<-14}
    if(temp.ID == "Hastigerinidae"){col.code[i]<-15}
    if(temp.ID == "Hedbergellidae"){col.code[i]<-16}
    if(temp.ID == "Heterohelicidae"){col.code[i]<-19}
    if(temp.ID == "Planomalinidae"){col.code[i]<-21}
    if(temp.ID == "Rotaliporidae"){col.code[i]<-22}
    if(temp.ID == "Rugoglobigerinidae"){col.code[i]<-23}
    if(temp.ID == "Schackoinidae"){col.code[i]<-27}
    if(temp.ID == "Truncorotaloididae"){col.code[i]<-31}
    if(temp.ID == "UNKNOWN"){col.code[i]<-32}
    if(temp.ID == "Paraticinellinae"){col.code[i]<-33}
  }
  col.key<-cbind(unique(family.ID),col.code) #col.key is now a 2-col obj with family (c1) and color.index (c2)
  for(i in 1:nrow(col.key)){
    col.key[i,2]<-col.temp[as.numeric(col.key[i,2])]
  }
  1->counter
  for(i in unique(family.ID)){ #This portion takes colkey and subs in the color value from RColorBrewer
    fam.col[which(family.ID == i)]<-col.key[which(col.key[,1] == i),2]
    counter+1->counter
  }
  plot(0,0,type='n',
       xlim=c(max(as.vector(origin),na.rm=T),-{spacer*length(unique(family.ID))*.9}),
       ylim=c(min(var.A,na.rm=T),max(var.A,na.rm=T)),
       ylab=paste(y.axis.label),
       xlab='Time (Ma)'
  )
  1->counter
  aggregate(var.A,by=list(family.ID),FUN=median,na.rm=T)->a.order
  order(a.order[,2])->a.order.1
  a.order[a.order.1,1]->a.order#setting an order for the vioplots
  for(i in a.order){
    if(all(is.na(var.A[which(family.ID == i)])) != T){
      segments(as.vector(origin[which(family.ID == i)]),
               var.A[which(family.ID == i)],
               as.vector(extin[which(family.ID == i)]),
               var.A[which(family.ID == i)],
               col=col.key[which(col.key[,1] == i),2],
               lwd=2
      )
      vioplot(na.omit(var.A[which(family.ID == i)]),
              horizontal=F,
              at=-{counter-0.5}*spacer,
              add=T,
              border=F,
              col=col.key[which(col.key[,1] == i),2],
              wex=spacer,
              na.rm=T)
      text(-{counter-0.5}*spacer,
           -10,
           labels=i,
           col=col.key[which(col.key[,1] == i),2],
           srt=90
           
      )
    }
    counter+1->counter
  }
}
```
```{r single morphology plot}
#basic plot
plot(0,0,
     type='n',
     xlim=c(170,0),
     ylim=c(min(foram.dataframe$fcarea,na.rm=T),
            max(foram.dataframe$fcarea,na.rm=T)),
     xlab='Time (Ma)',
     ylab="fcarea")
segments(foram.dataframe$origin,
         foram.dataframe$fcarea,
         foram.dataframe$extin,
         foram.dataframe$fcarea)
```
```{r making test complexity index.1}
inc<-c(#"w",
  #"mtheta",
  #"lw",
  #"lh",
  #"ic1",
  #"ic2",
  #"ic3",
  "fcirc",
  "t",
  "numcham",
  "expans",
  #"height",
  #"length",
  #"fcangle",
  "area",
  "fcarea",
  "clava",
  #"chamwl",
  "keel",
  #"bidors",
  #"biven",
  "biconvex",
  "lobe"
  #"double"
  #"depth"
)
pairs(morph[,inc],pch=16,cex=.4)  
```
```{r making test complexity.2}
which(morph$species == "holmdelensis")->h
which(morph$species == "monmouthensis")->m
build.tci<-morph
for(i in inc){
  {morph[h,i]+morph[m,i]}/2->simple.value
  build.tci[,i]<-morph[,i]-simple.value
}
plot(0,0,
     type='n',
     xlim=c(170,0),
     ylim=c(min(build.tci$fcarea,na.rm=T),
            max(build.tci$fcarea,na.rm=T)),
     xlab='Time (Ma)',
     ylab="fcarea minus simple")
segments(build.tci$origin,
         build.tci$fcarea,
         build.tci$extin,
         build.tci$fcarea)
```

So here's a plot showing the a issue with picking one end. there are changes in the final chamber area/test area that are highly different from Meso->Ceno. Which is odd, but it's probably the orbulinids and others like it. Either way, our assigned simple bugs are smack in the middle. Neat how it drifts one direction and the next.

```{r building test complexity index.3}
for(i in inc){
  abs(build.tci[,i])->build.tci[,i]#taking the absolute value
  decostand(build.tci[,i],method='range',na.rm=TRUE)->build.tci[,i]
}
test.complex.index<-NA
for(i in 1:length(build.tci$species)){
  sum(build.tci[i,inc])->test.complex.index[i]
}
plot(0,0,
     type='n',
     xlim=c(170,0),
     ylim=c(min(test.complex.index,na.rm=T),
            max(test.complex.index,na.rm=T)),
     xlab='Time (Ma)',
     ylab="test.complex.index")
segments(build.tci$origin,
         test.complex.index,
         build.tci$extin,
         test.complex.index)
```

Yep, still got roughly the same expression. 2 survivor bugs are the ones at *almost* 0 down there. There's two if you look really really closely

```{r TCI family}
morph$family->family.ID
range.line.plot(test.complex.index,
                "Test Complexity",
                family.ID,
                morph$origin,
                morph$extin
                )
```


Here's the breakdown of the families. 

```{r inc mean median var}
plot(0,0,
     type='n',
     xlim=c(170,0),
     ylim=c(min(test.complex.index,na.rm=T),
            max(test.complex.index,na.rm=T)),
     xlab='Time (Ma)',
     ylab="test.complex.index")
segments(build.tci$origin,
         test.complex.index,
         build.tci$extin,
         test.complex.index)
lines(time.mean(test.complex.index,morph$origin,morph$extin,.25),lwd=2)
lines(time.median(test.complex.index,morph$origin,morph$extin,.25),lwd=2,col="red")
  time.var(test.complex.index,morph$origin,morph$extin,.25)->temp
  lines(temp[,1],temp[,2],lwd=2,col='blue')
```





```{r mean per div}
time.mean(test.complex.index,morph$origin,morph$extin,.25)[,2]/time.div(test.complex.index,morph$origin,morph$extin,.25)[,2]->meanperdiv
plot(0,0,
     type='n',
     xlim=c(170,0),
     ylim=c(0,
            .2),
     xlab='Time (Ma)',
     ylab="Mean / Diversity")
axis(side=1,at=seq(0,170,by=10))
lines(time.mean(test.complex.index,morph$origin,morph$extin,.25)[,1],meanperdiv,lwd=2)
abline(v=c(23,55,34))
abline(v=OAEs.dataframe$oae.start)
```


OK, back to the same story. Higher mean TCI scores during recovery as morphology is colonized (adaptive radiation).